---
title: "spatial_marathon"
author: "Lelia Rønnow Corfitsen"
date: "2025-05-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load packages 
```{r}
library(gdistance)
library(sp)
library(raster)
library(sf)
```

## load and prepare the data 
```{r}
# load the DEM
dem <- raster("attica_dem.tif")
```

```{r}
# define coordinates for marathon, athens, and sparta
points <- data.frame(
  name = c("Marathon", "Athens", "Sparta"),
  lon = c(24.017, 23.73, 22.43),
  lat = c(38.15, 37.98, 37.07)
)
```

```{r}
# create sf object
coordinates <- st_as_sf(points, coords = c("lon", "lat"), crs = 4326)

# project the coordinates to the same CRS as the DEM
points_projected <- st_transform(coordinates, crs = st_crs(dem))
```

## Distance calculations 
```{r}
# get the geodetic distance between Marathon and Athens
marathon_athens <- st_distance(coordinates[1, ], coordinates[2, ])

# get the geodetic distance between Athens and Sparta
athens_sparta <- st_distance(coordinates[2, ], coordinates[3, ])
```

```{r}
# get the euclidean distance between Marathon and Athens
euc_dist <- st_distance(points_projected[1, ], points_projected[2, ])

# get the euclidean distance between Athens and Sparta
euc_dist_sparta <- st_distance(points_projected[2, ], points_projected[3, ])
```

```{r}
# print the distances in km
print(paste("Euclidean distance (2D) Marathon -> Athens (km):", euc_dist / 1000))
print(paste("Euclidean distance (2D) Athens -> Sparta (km):", euc_dist_sparta / 1000))
print(paste("Geodetic distance (3D) Marathon -> Athens (km):", marathon_athens / 1000))
print(paste("Geodetic distance (3D) Athens -> Sparta (km):", athens_sparta / 1000))
```

# Tobler's Hiking Function
The output of Tobler's hiking function is the walking velocity in km/hr. When using the function, the slope of the terraine is calculated and converted to a walking velocity. The function is defined as:
```{r}
tobler <- function(slope) {
  6 * exp(-3.5 * abs(slope + 0.05))
}
```

# Analysis of the terrain
```{r}
# reduce resolution of the DEM
raster_lowres <- aggregate(dem, fact = 10, fun = mean)
```

```{r}
# transition matrix
slope <- terrain(raster_lowres, opt = "slope", unit = "radians")
transition <- transition(tobler(slope), transitionFunction = mean, directions = 8)
trC <- geoCorrection(transition, type = "c")
```

# Least-cost path analysis
```{r}
# define the cost path function
cost_path <- function(from, to, trC) {
  cost <- costDistance(trC, from, to)
  path <- shortestPath(trC, from, to, output = "SpatialLines")
  list(cost = cost, path = path)
}
```

```{r}
# prepare SpatialPoints
points <- as(points_projected, "Spatial")
```

```{r}
# Calculate the paths
marathon_to_athens <- cost_path(points[1, ], points[2, ], trC)
sparta_to_athens <- cost_path(points[3, ], points[2, ], trC)
```

# Visualisations
```{r}
# plot the least cost paths
plot(raster_lowres, main = " Least-cost path from Marathon and Sparta to Athens")
lines(marathon_to_athens$path, col = "blue", lwd = 2)
lines(sparta_to_athens$path, col = "red", lwd = 2)
points(points, pch = 16, col = "black")
```

# Output results 
```{r}
# estimated travel time
print(paste("Estimated time (hours) Marathon -> Athens:", marathon_to_athens$cost))
print(paste("Estimated time (hours) Sparta -> Athens:", sparta_to_athens$cost))

```

```{r}
# Convert line path to projected CRS to get accurate distance
marathon_path_length <- SpatialLinesLengths(marathon_to_athens$path) / 1000  # in km
sparta_path_length <- SpatialLinesLengths(sparta_to_athens$path) / 1000      # in km

print(paste("Terrain-aware distance Marathon -> Athens (km):", marathon_path_length))
print(paste("Terrain-aware distance Sparta -> Athens (km):", sparta_path_length))
```

```{r}
# straight-line distance
distance_athens_sparta <- spDistsN1(points[2, ], points[3, ], longlat = TRUE)
print(paste("Estimated distance Athens -> Sparta (km):", distance_athens_sparta))
```

```{r}
# calorie burn estimation
kcal_burned_marathon <- marathon_to_athens$cost * 0.5
print(paste("Estimated kcal burned Marathon -> Athens:", kcal_burned_marathon))
```

```{r}
# get and print estimated kcal burned from sparta to athens
kcal_burned_sparta <- sparta_to_athens$cost * 0.5
print(paste("Estimated kcal burned Sparta -> Athens:", kcal_burned_sparta))
```




















